---
title: "Fitting Exercise Joe Dainis"
---
```{r}
#ensure packages are installed
#load the necessary libraries
library(dplyr)
library(purrr)
library(lubridate)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tidyr)
library(gtsummary)

#Load the data
fittingdata <- read.csv("fittingexercisedata.csv")

#Summarize the data
summary(fittingdata)
```
```{r}
#Plot DV as a function of time stratified by dose
fittingdata %>%
  filter(EVID == 0, MDV == 0) %>%
  ggplot(aes(x = TIME,
             y = DV,
             group = ID)) +
  geom_line(alpha = 0.6, color = "steelblue") +
  geom_point(size = 1.5) +
  facet_wrap(~ DOSE) +
  labs(
    x = "Time",
    y = "DV"
  ) +
  theme_bw()
```

### Conclusion: The three doses of mavoglurant appear to have similar trends of DV over time.

```{r}
#Keep only entries where patients received one dose
fittingdata_occ1 <- fittingdata %>%
  filter(OCC == 1)

#sumamrize data where patients only received one dose
summary(fittingdata_occ1)

#Compute the sum of DV for each individual after excluding observations where TIME = 0
sum_fittingdata <- fittingdata_occ1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(
    Y = sum(DV, na.rm = TRUE),
    .groups = "drop"
  )

#Make a dataframe which only contains data where TIME = 0
time0_fittingdata <- fittingdata_occ1 %>%
  filter(TIME == 0)

#Combine the two dataframes using the left_join() function
final_fittingdata <- time0_fittingdata %>%
  left_join(sum_fittingdata, by = "ID")

#summarize combined data table
summary(final_fittingdata)

#Clean the data to convert RACE and SEX to factor variables keeps the following variables: Y, DOSE, AGE, SEX, RACE, WT, HT
analysis_fittingdata <- final_fittingdata %>%
  mutate(
    RACE = factor(RACE),
    SEX  = factor(SEX)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

#generate a summary table for the data, including mean and standard deviation of the weight, height, age, and Y variable and splits of how many were in each dose, sex and race group
analysis_fittingdata %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    )
  )

#Save analysis_fittingdata as an RDS file
saveRDS(analysis_fittingdata, file = "analysis_fittingdata.rds")
```

### Conclusion: This table shows that many of the participants in the study received either a dose of 25 or 50, with a smaller representation for the medium dosage. There also appears to be a major skew towards one sex, and a large skew towards one race.

```{r}
#summarize the data by the three dosage groups
analysis_fittingdata %>%
  group_by(DOSE) %>%
  summarise(
    n = n(),
    mean_Y = mean(Y),
    sd_Y = sd(Y),
    median_Y = median(Y),
    min_Y = min(Y),
    max_Y = max(Y)
  )

#Generate preliminary relationship graphs between-
##Age and Y
ggplot(analysis_fittingdata, aes(AGE, Y)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

### Conclusion: No correlation between Age and Outcome
```{r}
##Weight and Y
ggplot(analysis_fittingdata, aes(WT, Y)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

### Conclusion: Slight negative correlation between Weight and Outcome
```{r}
##Height and Y
ggplot(analysis_fittingdata, aes(HT, Y)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

### Conclusion: Slight negative correlation between Height and Outcome

```{r}
##Sex and Y
ggplot(analysis_fittingdata, aes(SEX, Y)) +
  geom_boxplot() +
  theme_bw()
```

### Conclusion: No difference noted between Sex and Outcome
```{r}
##Race and Y
ggplot(analysis_fittingdata, aes(RACE, Y)) +
  geom_boxplot() +
  theme_bw()
```

### Conclusion: No difference noted between Race and Outcome
```{r}
##Dose and Y
ggplot(analysis_fittingdata, aes(factor(DOSE), Y)) +
  geom_boxplot() +
  theme_bw()
```

### Conclusion: An increasing dose dramatically correlates with an increase in Outcome Y
```{r}
#Plot the distributions of the variables

##Histogram of the outcome, Y, to see distribution of outcome in the data
ggplot(analysis_fittingdata, aes(Y)) +
  geom_histogram(bins = 30) +
  theme_bw()

ggplot(analysis_fittingdata, aes(Y)) +
  geom_density() +
  theme_bw()
```

### Conclusion: Left skew to the distribution, with greater representation of Y outcomes smaller than 3000.
```{r}
##Distribution of age in the dataset
ggplot(analysis_fittingdata, aes(AGE)) +
  geom_histogram(bins = 20) +
  theme_bw()
```

### Conclusion: Possibly bimodal distribution due to the two peaks in ages seen
```{r}
##Distribution of weight in the dataset
ggplot(analysis_fittingdata, aes(WT)) +
  geom_histogram(bins = 20) +
  theme_bw()
```

### Conclusion: Normal Distribution of weight
```{r}
##Distribution of height in the dataset
ggplot(analysis_fittingdata, aes(HT)) +
  geom_histogram(bins = 20) +
  theme_bw()
```

### Conclusion: Normal Distribution of height
```{r}
#Generate some correlation/pair plots to see correlations in the data
##Make a visual correlation plot
###Load the corrplot library
library(corrplot)

cor_fittingdata <- analysis_fittingdata %>%
  select(Y, DOSE, AGE, WT, HT) %>%
  cor()

corrplot(cor_fittingdata, method = "color")

##Make a pair plot
###Load the corrplot library
library(GGally)

analysis_fittingdata %>%
  select(Y, DOSE, AGE, WT, HT) %>%
  ggpairs()
```

# Main Takeaways
1. There appears to be a significant correlation between the dose of mavoglurant and the pharmokinetic Y outcome, which implies that the drug is having some sort of effect on the outcome.
2. There also appears to be a slight correlation between weight and Y outcome, with higher weights correlating with a smaller Y.
3. The distribution of Y outcome values is slightly skewed to the left, with a slightly higher representation of smaller Y values than larger Y values. A majority of the Y values are smaller than 3000.

## Run Linear Models
```{r}
#Load necessary libraries
library(yardstick)
library(pROC)

#Fit a linear model and use the DOSE variable to predict the Y outcome variable

##Run the linear model
lm_dose <- lm(Y ~ DOSE, data = analysis_fittingdata)

##Predict the outcome based on the dose
pred_lm_dose <- analysis_fittingdata %>%
  mutate(pred = predict(lm_dose, newdata = analysis_fittingdata))

##Run the metrics to see how well the model fit
rmse_lm_dose <- rmse(pred_lm_dose, truth = Y, estimate = pred)
rsq_lm_dose  <- rsq(pred_lm_dose, truth = Y, estimate = pred)

#Display the RMSE and RSQ metrics
print(rmse_lm_dose)
print(rsq_lm_dose)
```

### Conclusion: The linear model using DOSE as a predictor of the outcome is weak but not unusable.

```{r}
#Fit a linear model and use all variables to predict the Y outcome variable

##Run the linear model using all variables
lm_all <- lm(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = analysis_fittingdata)

##Predict the outcome based on all variables
pred_lm_all <- analysis_fittingdata %>%
  mutate(pred = predict(lm_all, newdata = analysis_fittingdata))

##Run the metrics to see how well the model fit
rmse_lm_all <- rmse(pred_lm_all, truth = Y, estimate = pred)
rsq_lm_all  <- rsq(pred_lm_all, truth = Y, estimate = pred)

#Display the RMSE and RSQ metrics
print(rmse_lm_all)
print(rsq_lm_all)
```

### Conclusion: The linear model using all variables as a predictor of the outcome is weak but not unusable.

```{r}
#Fit a logsitic model and use DOSE to predict the SEX variable

##Run the logsitic model using the SEX and DOSE variables
logit_dose <- glm(SEX ~ DOSE,
                  data = analysis_fittingdata,
                  family = binomial)

##Predict the outcome based on the DOSE variable
pred_logit_dose <- analysis_fittingdata %>%
  mutate(
    prob = predict(logit_dose, type = "response"),
    pred_class = factor(ifelse(prob > 0.5, 1, 0),
                        levels = levels(SEX))
  )

##Generate the accuracy metrics to see how well the model fit
acc_logit_dose <- accuracy(pred_logit_dose,
                           truth = SEX,
                           estimate = pred_class)

print(acc_logit_dose)

##Generate the ROC metrics to see how well the model fit
roc_dose <- roc(analysis_fittingdata$SEX, pred_logit_dose$prob)
print(auc(roc_dose))
```

### Conclusion: The logsitic model using DOSE as a predictor of SEX is not usable.

```{r}
#Fit a logsitic model and use all variables to predict the SEX variable

##Run the logsitic model using the SEX and DOSE variables
logit_all <- glm(SEX ~ DOSE + AGE + RACE + WT + HT,
                  data = analysis_fittingdata,
                  family = binomial)

##Predict the outcome based on the DOSE variable
pred_logit_all <- analysis_fittingdata %>%
  mutate(
    prob = predict(logit_all, type = "response"),
    pred_class = factor(ifelse(prob > 0.5, 1, 0),
                        levels = levels(SEX))
  )

##Generate the accuracy metrics to see how well the model fit
acc_logit_all <- accuracy(pred_logit_all,
                           truth = SEX,
                           estimate = pred_class)

print(acc_logit_all)

##Generate the ROC metrics to see how well the model fit
roc_all <- roc(analysis_fittingdata$SEX, pred_logit_all$prob)
print(auc(roc_all))
```

### Conclusion: The logsitic model using all variables as a predictor of SEX is not usable.

# What Does This All Mean?

1. The first conclusion we can come to is that using DOSE as a predictor of the Y outcome variable is quite weak. Our RMSE metric was 666 and the R-squared was 0.516, indicating that the model is not total and utter garbage, but that using dose alone to predict the outcome is not reliable.
2. We see that including all of the variables into the linear regression model improved the metrics (smaller RMSE to 591, larger R-squared to 0.619), but still, all of the variables are not enough to strongly predict the outcome. This matches what I would expect with complicated drug data, with many variables likely impacting the outcome. This could also indicate trialing out higher doses (if possible) to see if this predictive value could get any stronger.
3. SEX is a categorical variable that cannot be predicted by these variables (which we knew coming into this analysis).

# Module 10: Model Improvement

## Subset Fitting Dataset and Make Training/Testing Datasets
```{r}
#Load libraries
library(tidymodels)

#Subset data to remove RACE variable and keep 6 other variables
subset_fittingdata <- analysis_fittingdata %>%
  select(Y, DOSE, AGE, SEX, WT, HT)

#Set seed for reproducibility
set.seed(1234)

#Split the data so that 75% is for the training dataset and 25% is for the testing dataset
subset_fittingdata_split <- initial_split(subset_fittingdata, prop = 0.75)

##Make training dataset
fittingdata_training <- training(subset_fittingdata_split)

##Make testing dataset
fittingdata_testing <- testing(subset_fittingdata_split)
```

## Fit Linear Models to the training data
```{r}
#Model 1: Fit a linear model to predict the outcome variable Y using the DOSE variable as the predictor
##Set engine to a linear model
lm_model <- linear_reg() %>%
  set_engine("lm")

##Run linear model using DOSE as a predictor of Y outcome
model_dose <- workflow() %>%
  add_model(lm_model) %>%
  add_formula(Y ~ DOSE)

##Fit linear model of dose to the training data
fit_dose <- fit(model_dose, data = fittingdata_training)

#Model 2: Fit a linear model for Y but this time use all variables as the predictor

##Run linear model using All Variables as a predictor of Y outcome
model_all <- workflow() %>%
  add_model(lm_model) %>%
  add_formula(Y ~ DOSE + AGE + SEX + WT + HT)

##Fit linear model of all variables to the training data
fit_all <- fit(model_all, data = fittingdata_training)
```

## Model Performance Assessment 1
```{r}
#Compute Predictions for the two models
##Dose Linear Model Prediction
pred_dose <- predict(fit_dose, fittingdata_training) %>%
  bind_cols(fittingdata_training)

##All Variables Linear Model Prediction
pred_all <- predict(fit_all, fittingdata_training) %>%
  bind_cols(fittingdata_training)

#Computer RMSE for the two models
##RMSE for Dose Linear Model;
rmse_dose <- rmse(pred_dose, truth = Y, estimate = .pred)

##RMSE for All Variables Linear Model
rmse_all  <- rmse(pred_all, truth = Y, estimate = .pred)

print(rmse_dose)
print(rmse_all) 

#Compute the RMSE of a Null Model
##Calculate the mean of the Y variable from the fitting data
mean_y <- mean(fittingdata_training$Y)

##Calculate the RMSE for the Null Model
rmse_null <- sqrt(mean((fittingdata_training$Y - mean_y)^2))
print(rmse_null)
```

## Conclusions and Comparisons of the Models using Model Assessment 1
The RMSE values predicted/calculated for the models are as follows:
Null Model: 948.3526
Dose Model: 703
All Variables Model: 627
Based on these RMSE values, the All Variables Model is performing the best, as it has the smallest RMSE value. This also shows that although the Dose Model is not the best model compared to all of the variables as predictors, it is a major improvement over the Null Model.

## Model Performance Assessment 2
```{r}
#Load libraries
library(tidymodels)

#Run a 10-fold cross validation to assess model performance

##Reset the seed
set.seed(1234)

##Create folds
folds <- vfold_cv(fittingdata_training, v = 10)

##Run 10-fold cross validation for DOSE model
cv_dose <- fit_resamples(
  fit_dose,
  resamples = folds,
  metrics = metric_set(rmse)
)

##Run 10-fold cross validation for all variables model
cv_all <- fit_resamples(
  fit_all,
  resamples = folds,
  metrics = metric_set(rmse)
)

##Collect results
collect_metrics(cv_dose)
collect_metrics(cv_all)
```

## Conclusions and Comparisons of the Models using Model Assessment 2 and mean RMSEs
The mean RMSE values predicted/calculated for the models are as follows:
Null Model: 948.3526
Dose Model: 691 +/- 67.5 (RMSE standard error)
All Variables Model: 646 +/- 64.8 (RMSE standard error)
It appears there are slight changes to how the models performed using this model assessment method. The All Variables Model still performs the best compared to the Null and Dose Models, so using all variables as predictors still remains the best model. However, the RMSE values shifted, with the Dose Model having a smaller RMSE using this second model assessment and the All Variables Model having a larger RMSE. They both appear to have a very similar standard error. This model assessment shows that although the All Variables Model is still the best model, it is much closer than previously, and that the Dose variable is driving much of the prediction.

## Re-run Model Assessment with a Different Seed
```{r}
#Load libraries
library(tidymodels)

#Re-run a 10-fold cross validation to assess model performance

##Reset the seed
set.seed(61940)

##Create folds
folds <- vfold_cv(fittingdata_training, v = 10)

##Run 10-fold cross validation for DOSE model
cv_dose2 <- fit_resamples(
  fit_dose,
  resamples = folds,
  metrics = metric_set(rmse)
)

##Run 10-fold cross validation for all variables model
cv_all2 <- fit_resamples(
  fit_all,
  resamples = folds,
  metrics = metric_set(rmse)
)

##Collect results
collect_metrics(cv_dose2)
collect_metrics(cv_all2)
```

## Conclusions and Comparisons of the Models using Model Assessment 2 Re-run with New Seed
The RMSE values predicted/calculated for the models are as follows:
Null Model: 948.3526
Dose Model: 677 +/- 73.5 (RMSE standard error)
All Variables Model: 621 +/- 74.4 (RMSE standard error)
Increasing the seed lead to a better mean RMSE but higher standard error.

# This section added by Liang Lai

## Model Predictions vs Observed Values
```{r}
# Combine predictions from all 3 models into one data frame
plot_preds <- bind_rows(
  fittingdata_training %>%
    select(Y) %>%
    mutate(.pred = mean_y,
           model = "Null Model"),
  pred_dose %>%
    select(Y, .pred) %>%
    mutate(model = "Model 1 (DOSE only)"),
  pred_all %>%
    select(Y, .pred) %>%
    mutate(model = "Model 2 (All predictors)")
)

# Plot observed vs predicted
ggplot(plot_preds, aes(x = Y, y = .pred, 
                       color = model, shape = model)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(0, 5000) + ylim(0, 5000) +
  labs(
    title = "Observed vs Predicted Values",
    x = "Observed",
    y = "Predicted",
    color = "Model",
    shape = "Model"
  ) +
  theme_bw()
```

The null model predicts a single horizontal line (the mean), Model 1 produces three horizontal lines corresponding to the three dose levels, and Model 2 best follows the 45-degree line but still shows notable scatter, especially underestimating higher observed values.


## Residual Plot for Model 2
```{r}
# Compute residuals for Model 2
resid_data <- pred_all %>%
  mutate(residual = .pred - Y)

# Plot predicted vs residuals
ggplot(resid_data, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(-2500, 2500) +
  labs(
    title = "Predicted vs Residuals (Model 2)",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_bw()
```

The residual plot shows considerable scatter and a noticeable spreading trend as predicted values increase. This suggests that the model's errors are not constant across the range of predictions, indicating potential heteroscedasticity and that the model may be underfitting the data, especially at higher predicted values.

## Bootstrap Uncertainty for Model 2
```{r}
# Reset seed
set.seed(1234)

# Create 100 bootstraps from training data
dat_bs <- bootstraps(fittingdata_training, times = 100)

# Store predictions in a matrix
pred_bs <- matrix(NA,
                  nrow = 100,
                  ncol = nrow(fittingdata_training))

# Loop over bootstrap samples and fit model to each sample
for (i in 1:100) {
  dat_sample <- rsample::analysis(dat_bs$splits[[i]])
  fit_i <- fit(model_all, data = dat_sample)
  pred_bs[i, ] <- predict(fit_i,
                          new_data = fittingdata_training)$.pred
}

# Compute median and 89% confidence intervals across bootstrap samples
preds <- pred_bs |>
  apply(2, quantile, c(0.055, 0.5, 0.945)) |> t()

# Combine original predictions, bootstrap median and CI into one data frame
boot_results <- tibble(
  observed = fittingdata_training$Y,
  original = pred_all$.pred,
  lower    = preds[, 1],
  median   = preds[, 2],
  upper    = preds[, 3]
)

# Plot observed vs predicted with bootstrap uncertainty
# Black = original predictions, Blue = bootstrap median, Red = 89% CI bounds
ggplot(boot_results, aes(x = observed)) +
  # Draw vertical lines connecting lower and upper CI bounds
  geom_segment(aes(xend = observed, y = lower, yend = upper),
               color = "lightblue", alpha = 0.5) +
  # Plot lower and upper CI bounds in red
  geom_point(aes(y = lower), color = "red", alpha = 0.5) +
  geom_point(aes(y = upper), color = "red", alpha = 0.5) +
  # Plot bootstrap median in blue
  geom_point(aes(y = median), color = "blue", alpha = 0.5) +
  # Plot original predictions in black (on top)
  geom_point(aes(y = original), color = "black") +
  # Add 45-degree reference line
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(0, 5000) + ylim(0, 5000) +
  labs(
    title = "Bootstrap Predictions with Uncertainty (Model 2)",
    x = "Observed",
    y = "Predicted"
  ) +
  theme_bw()
```

The plot displays observed values against predicted values for Model 2, including the point estimates (black) and the 89% bootstrap confidence intervals (red/blue). While the predictions generally follow the 45-degree reference line, there is substantial scatter, indicating a large amount of unexplained variance. Notably, for higher observed values (e.g., >3000), the predictions consistently fall below the 45-degree line, revealing a systematic underestimation by the model. The bootstrap confidence intervals are relatively narrow and stable across the range of values; however, for the higher observed values, these intervals frequently fail to capture the true observed value (the 45-degree line).

# This final part 3 added by Joe Dainis

## Fit of Model 2 using the test data
```{r}
#Predict on the saved testing data using Model 2
predicttest_all <- predict(fit_all, new_data = fittingdata_testing) %>%
  bind_cols(fittingdata_testing)

#Compute performance metrics on saved testing data
rmse_test_all <- rmse(predicttest_all, truth = Y, estimate = .pred)
rsq_test_all  <- rsq(predicttest_all, truth = Y, estimate = .pred)

print(rmse_test_all)
print(rsq_test_all)
```

## Plot the predicted versus observed for the fitted model used on the test data

```{r}
#Add labels to differentiate training data predictions and testing data predictions
train_plot <- pred_all %>%
  mutate(dataset = "Training")

test_plot <- predicttest_all %>%
  mutate(dataset = "Test")

#Combine the training and testing prediction datasets into one dataframe 
combined_plot <- bind_rows(train_plot, test_plot)

#Plot the training and testing prediction values versus the observed values in one plot
testvstraining_plot <- ggplot(combined_plot, aes(x = Y, y = .pred, color = dataset, shape = dataset)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(0, 5000) + ylim(0, 5000) +
  labs(
    title = "Observed vs Predicted (Training vs Test Data)",
    x = "Observed Y",
    y = "Predicted Y"
  ) +
  theme_bw()

testvstraining_plot
```

# Final Conclusions
1. Overall, we have validated that the model using dose as a predictor of Y outcome as well as the model using all variables as predictors both outperformed our Null model. (RMSE = 948 for Null reduced to RMSE = ~600-700 depending on model used and assessment).
2. Dose accounts for a majority of the predicting ability, as using all variables sees an additional but very slight improvement in predicting ability compared to using dose alone. We also see that the model applied to the testing data intermeshes quite nicely with the training data, indicating that this model was successful in its development.
3. Overall, the results do make sense to me. It would make sense that a clinically approved drug in Mavoglurant used at higher doses would have an increasing detectable concentration of the drug. The dose model might be useful for assessing general trends, but it is not a perfect predictor and it would likely not be useful to use to predict on an individual case-by-case basis.
4. Finally, for the All Variables model, as mentioned before it does see a slight improvement in predictive power, but still since it is only a slight improvement I would not use this on an individual case basis to predict an outcome. It does make sense that using all of the available variables would slightly improve predictive capability of the outcome, but this also proves that the main variable driving the predictive power of the model is dose, which makes sense for this dataset studying the pharmokinetics of the drug.